### 2021-12-09 text/jsm/jsm_paper.tex
add jsm text

\documentclass[12pt]{article}
 \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{array}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother
\setcounter{tocdepth}{2}
\usepackage{xcolor}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\usepackage{float}
\allowdisplaybreaks
\usepackage[caption = false]{subfig}
\usepackage{/Users/timbarry/optionFiles/mymacros}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}
\noindent
Tim, Gene, Kathryn
\begin{center} 
\textbf{A class of exponential family measurement error models, with application to CRISPR genome editing and single-cell sequencing}
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression in challenging problem settings. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism using a new class of errors-in-variables (or measurement error) models. We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$. We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$. The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x , y)$ only. In the context of the biological application, $x^*$, $y$, and $x$ are CRISPR perturbations, gene expressions, and guide RNA counts, respectively.

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method. Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets. % In highly noisy settings ; by contrast, in less noisy settings, GLM-EIV serves as a principled and reproducible threshold selection procedure.

@@

\subsection{Related work}

Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses. Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models negative binomially-distributed gene expression counts while controlling for technical factors. We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}. Gr\"{u}n \& Leisch derived a method for estimation and inference in a $k$-component mixture of GLMs. While we prefer to view GLM-EIV as a generalized errors-in-variables method,  the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities. Ibrahim proposed a procedure for fitting GLMs in the presence of missing-at-random covariates. Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate. Thus, while Ibrahim and Gr\"{u}n \& Leisch are helpful references, our estimation and inference tasks are more complex than theirs.

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}. These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.

@@

\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design. For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed. Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene. \textbf{b,} DAG representing all variables in the system. The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression. The target of estimation is the effect of the perturbation on gene expression. \textbf{c,} Schematic illustrating the ``background read'' phenomenon. Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom). The target of estimation is the change in mean gene expression in response to the perturbation (top). \textbf{d}, Example data on four cells for a given perturbation-gene pair. Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
	\label{analysis_challenges}
\end{figure}

@@

The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current discussion). Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}. These noncoding variants are thought to contribute to disease by modulating the expression one or more disease-relevant genes. Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results. A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.

The most promising biotechnology for solving this challenge are high MOI single-cell CRISPR screens. High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that represses a targeted region of the genome -- with single-cell sequencing. The experimental protocol is as follows. First, the scientist develops a library of several hundred to several thousand CRISPRi perturbations, each designed to target a candidate enhancer for repression. The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to these cells. The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations. Conversely, a given perturbation enters about 0.1-2\% of cells.

@@

In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ and Xie et al.\ in 2019. Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells. Gasperini additionally designed 381 positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.

\section{Thresholding method}
We study the thresholding method from empirical and theoretical perspectives, highlighting several limitations of the approach. Gasperini and Xie both imputed perturbation assignments onto the cells via thresholding, but they carried out subsequent differential expression analyses in different ways: Gasperini used negative binomial regression, whereas Xie applied nonparametric tests of independence. We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural approach.

Let $n \in \N$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
$$m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i),$$
\begin{equation}\label{thresh_glm}
\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
\end{equation}
where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.

\item[3.]  Fit a GLM to obtain estimates of the parameters. Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}) yields
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ The term $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration. The target of inference, $\beta^m_1$, is the log-transformed fold change in gene expression in response to the perturbation, controlling for the technical factors. Fold change (i.e., $\exp(\beta^m_1)$) is the ratio of mean gene expression in perturbed cells to unperturbed cells: $\exp(\beta^m_1) = 1$ indicates no change in expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).

@@

We examined the behavior of the thresholding method on real data and uncovered attenuation bias and bias-variance tradeoff effects. We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong decrease in expression.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b). The selected threshold substantially impacts the results. \textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5. The latter threshold yields more confident estimates. \textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data (0 counts not shown). The gRNA data do not appear to imply an obvious threshold selection strategy.}
	\label{thresholding_empirical}
\end{figure}
\newpage

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
	\caption{\textbf{Theoretical challenges of thresholded regression.} \textbf{a,} Asymptotic relative bias versus threshold for different values of $\beta^g_1$. The bias function is highly nonconvex and strictly nonzero. Vertical blue lines, Bayes-optimal decision boundaries. Across all panels, $\beta^g_0 = 0$ and $\pi = 1/2$. \textbf{b,} Asymptotic relative bias versus $\pi$ when the threshold is set to a large number. The two quantities coincide exactly. \textbf{c,} Bias-variance decomposition for thresholding method in no-intercept model. Bias decreases and variance increases as the threshold tends to infinity. $\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$.}
	\label{thresholding_theoretical}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/main_text_sim/plot}
	\caption{\textbf{Simulation study}. Columns correspond to distributions (Poisson, NB with known $\theta$, NB with estimated $\theta$), and rows correspond to metrics (bias, MSE, coverage, CI width, and time). Methods are shown in different colors; GLM-EIV (red) is masked by accelerated GLM-EIV (blue) in several panels. GLM-EIV demonstrated superior statistical performance to the thresholding method on all metrics (rows 1-4). Accelerated GLM-EIV had substantially lower computational cost than ``vanilla'' GLM-EIV (bottom row) despite demonstrating identical statistical performance (rows 1-4).}
	\label{main_text_sim}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
	\caption{\textbf{Data analysis}. \textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs. \textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination. Grey bands, 95\% CIs for the target of inference outputted by the methods. \textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination. Colored bands, pointwise 95\% CIs for population quantities. Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}\label{fig:real_data}
\end{figure}

@@

To our knowledge this is the first paper to examine single-cell CRISPR screens from a rigorous statistical perspective. We hope that this work can help introduce the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}. Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, an (in our view) tightly-integrated, user-friendly, and powerful set of computational tools for out-of-memory and massively distributed single-cell analysis.

We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data. For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infenction'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work). We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility data. A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}. We do not directly observe chromatin state; instead, we observe transposase-tagged DNA fragments, counts that serve as proxies for whether a given region of chromatin is open or closed. GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.

>>>>

\documentclass[11pt]{article}
\usepackage{setspace}
\doublespacing
\usepackage[margin=1in]{geometry}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}

\usepackage[
backend=biber,
style=numeric-comp,
maxcitenames=1,
maxbibnames=3,
backref=true,
doi=false,
isbn=false,
url=false,
eprint=false
]{biblatex}

\addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother
\setcounter{tocdepth}{2}
\usepackage{xcolor}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\usepackage{float}
\allowdisplaybreaks
\usepackage[caption = false]{subfig}
\usepackage{/Users/timbarry/optionFiles/mymacros}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}
\noindent
\begin{center} 
\textbf{A new class of exponential family measurement error models applied to single-cell CRISPR screens} \\
Timothy Barry, Kathryn Roeder, Eugene Katsevich
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression in challenging problem settings. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}

\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism using a new class of measurement error models. We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$ and vector of confounders $z$. We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$ and $z$. The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x, y, z)$ only. In the context of the biological application, $x^*$, $x$, $y$, and $z$ are CRISPR perturbations, guide RNA counts, gene expressions, and technical confounders, respectively.

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. % Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.

Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets. We find that in some settings, GLM-EIV outperforms thresholded regression by a considerable margin; in other settings the two methods work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.

@@

\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021}, improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction. We extend the classical errors-in-variables model in several directions, most importantly generalizing to response distributions and sources of measurement error that are exponential family-distributed.

The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}. Gr\"{u}n \& Leisch derived a method for estimation and inference in a $k$-component mixture of GLMs. While we prefer to view GLM-EIV as a generalized errors-in-variables method,  the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities. Ibrahim proposed a procedure for fitting GLMs in the presence of missing-at-random covariates. Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate. Thus, while Ibrahim and Gr\"{u}n \& Leisch are helpful references, our estimation and inference tasks are more complex than theirs. Other relevant references include Aigner \cite{Aigner1973} and Savoca \cite{Savoca2000}.

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we aim to account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Another difference is that we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is more challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

 %Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}. % These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.

@@

@@

The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current work). Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}. These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes. Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results. A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.

High MOI single-cell CRISPR screens are the most promising biotechnology for solving this challenge. High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that represses a targeted region of the genome -- with single-cell sequencing. The experimental protocol is as follows. First, the scientist develops a library of several hundred to several thousand CRISPRi perturbations, each designed to target a candidate enhancer for repression. The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to these cells. The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations. Conversely, a given perturbation enters about 0.1-2\% of cells.

@@

In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ \cite{Gasperini2019} and Xie et al.\ \cite{Xie2019a}. Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells. Gasperini additionally designed several hundred positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.

\section{Thresholding method}
We study thresholding from empirical and theoretical perspectives, highlighting several inherent difficulties of the approach. Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence. We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural method.

Let $n \in \N$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$ \hat{p}_i = 0 \textrm{ if } g_i < c; \quad \hat{p}_i = 1 \textrm{ if } g_i \geq c.$$
%$$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
\begin{equation}\label{thresh_glm}
m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i); \quad
\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
\end{equation}
where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.

\item[3.]  Fit a GLM to obtain an estimate and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression. Exponentiating both sides of (\ref{thresh_glm}) yields
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m,$$ where $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.

Fold change in this context is defined as the ratio of the mean expression of perturbed cells to the mean expression of unperturbed cells. The target of inference $\beta^m_1$ is the \textit{log} fold change in expression in response to the perturbation, controlling for the technical factors. Therefore, $\exp(\beta^m_1) = 1$ (eq., $\beta^m_1 = 0$) indicates no change in expression, while $\exp(\beta^m_1) > 1$ (eq., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (eq., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.

@@

We examined the behavior of the thresholding method on real data and uncovered attenuation bias and bias-variance tradeoff effects. We applied the thresholding method to analyze the set of 322 positive control perturbation-gene pairs in the Gasperini dataset. The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition. Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong decrease in expression.

@@

@@

@@

@@

@@

To our knowledge this is the first single-cell CRISPR screen paper oriented toward a statistical audience. We hope that this work helps to introduce the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}. Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a tightly-integrated, user-friendly, and powerful set of computational tools for large-scale single-cell analysis likely of interest to other researchers.

We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data. For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infection'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work). We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility assays. A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}. We do not directly observe chromatin state; instead, we observe tagged DNA fragments, counts that serve as proxies for whether a given region of chromatin is open or closed. GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets. 
\printbibliography
